---
title: "Near Redundancy and Practical non-identifiability: Lung cancer data"
author: '[F. Javier Rubio](https://sites.google.com/site/fjavierrubio67/), J.A. Espindola, and J.A. Montoya'
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
bibliography: references.bib
---

# Case study I: Lung cancer data
This document presents an analysis of Near Redundancy and Practical non-identifiability using the Lung cancer data set from the R package `survival`. For more information see @rubio:2019 and @rubio:2023.

See the GitHub repository: [Near Redundancy and Practical non-identifiability: Survival Models](https://github.com/FJRubio67/NRPNISurv)

**Acknolewdgement**: We thank Mr. Ching Choy for pointing out a bug in the bootstrap for the Hessian method. This bug produced a classification probability $\approx 1$, while the correct value is $\approx 0.7$, as shown in the section "Bootstrap for distance-based criteria and Hessian method''. The conclusions are similar for the new corrected values, but it is now clear that the Hessian method has a lower power in detecting inferential problems than that of the *ad hoc* distance-based methods, as expected. 

# Data preparation
```{r}
########################################################################
# Case study I: Lung cancer data set from the survival R package
########################################################################

# See README file for more information

rm(list=ls())

# Routines
# https://github.com/FJRubio67/NRPNISurv
source("routines.R")

# Required packages (from CRAN)
# install.packages('PackageName')
library(knitr)
library(numDeriv)
library(survival)
library(flexsurv)

# Required packages (from GitHub)
#library(devtools)
#install_github("FJRubio67/HazReg")
library(HazReg)

#----------------------------------------------------------------------------------------------
# Lung cancer data
#---------------------------------------------------------------------------------------------

# data set
attach(lung)
head(lung)

#******************
# Data preparation
#******************

# Removing a missing observation
dat <- lung[-14,]

# design matrix (hazard scale)
x <- as.matrix( cbind(scale(dat$age), dat$sex - 1, scale(dat$ph.ecog) ))

# sample size
n <- nrow(x)

# design matrix (time scale)
xt <- as.matrix(x[,1])

# vital status
status <- dat$status - 1

# Survival times
times <- dat$time/365.25

times  <- ifelse(times==0,1/365.25,times)
```


# Maximum likelihood estimation
```{r}
#----------------------------------------------------------------------------------------------
# Maximum likelihood estimation
# See : https://github.com/FJRubio67/HazReg
#----------------------------------------------------------------------------------------------

# PGW model with no covariates
OPTPGW0 <- GHMLE(init = c(0,0,0), times = times, status = status, 
               hstr = "baseline", dist = "PGW", method = "nlminb", maxit = 10000)

# PGW-PH model
OPTPGWPH <- GHMLE(init = c(OPTPGW0$OPT$par*0, rep(0,ncol(x))), times = times, status = status, 
           hstr = "PH", dist = "PGW", des = x, method = "nlminb", maxit = 10000)

# PGW-AFT model
OPTPGWAFT <- GHMLE(init = c(OPTPGW0$OPT$par,rep(0,ncol(x))), times = times, status = status, 
         hstr = "AFT", dist = "PGW", des = x, method = "nlminb", maxit = 10000)

# PGW-GH model
OPTPGWGH <- GHMLE(init = c(OPTPGW0$OPT$par, rep(0, ncol(xt)), rep(0,ncol(x))), times = times, status = status,                 
          hstr = "GH", dist = "PGW", des = x, des_t = xt, method = "nlminb", maxit = 10000)

# MLE PGW no covariates
MLE.PGW0 <- exp(OPTPGW0$OPT$par[1:3])
kable(MLE.PGW0, digits = 3)

# MLE PGW-PH structure
MLE.PGWPH <- c(exp(OPTPGWPH$OPT$par[1:3]), OPTPGWPH$OPT$par[-c(1:3)])
kable(MLE.PGWPH, digits = 3)

# MLE PGW-AFT structure
MLE.PGWAFT <- c(exp(OPTPGWAFT$OPT$par[1:3]), OPTPGWAFT$OPT$par[-c(1:3)])
kable(MLE.PGWAFT, digits = 3)

# MLE PGW-GH structure
MLE.PGWGH <- c(exp(OPTPGWGH$OPT$par[1:3]), OPTPGWGH$OPT$par[-c(1:3)])
kable(MLE.PGWGH, digits = 3)
```

# Model selection
```{r}
# AIC
AIC.PGW0 <- 2*OPTPGW0$OPT$objective + 2*length(MLE.PGW0)
AIC.PGWGH <- 2*OPTPGWGH$OPT$objective + 2*length(MLE.PGWGH)
AIC.PGWAFT <- 2*OPTPGWAFT$OPT$objective + 2*length(MLE.PGWAFT)
AIC.PGWPH <- 2*OPTPGWPH$OPT$objective + 2*length(MLE.PGWPH)

# Best model: PGW-PH
AICs <- c(AIC.PGW0, AIC.PGWPH, AIC.PGWAFT, AIC.PGWGH)
AICs
which.min(AICs)
```

# Detecting Near Redundancy
```{r}
######################################################################
# KL Divergence criterion
######################################################################
# KL Divergence
dKL <- minKL(MLE.PGWGH[1:3], c(1,1))$min.KL
dKL

# Redundancy constant
k = length(MLE.PGWGH)
M = 0.05
ne <- n - 0.5*sum(!status)

UKL <- 0.5*k*M*length(MLE.PGWGH)*log(ne)/ne

# Test
(dKL < UKL)



######################################################################
# Hellinger distance criterion
######################################################################

# Hellinger Distance
DH <- minHell(MLE.PGWGH[1:3], c(1,1))$min.dHell
DH

# Hellinger criterion
kappa <- 0.05
UH <- log( 1 - (1-2*kappa)^2)/(2*log(1-DH^2))

# Test
(ne < UH)


######################################################################
# Hessian Method
######################################################################

# Hessian/Eigenvalues
HESS.PGW  <- -hessian(OPTPGWGH$log_lik, x = OPTPGWGH$OPT$par)
  
eigen.val <- eigen(HESS.PGW)$values

ref.val <- abs(as.vector(eigen.val))/max(abs(as.vector(eigen.val)))

sev <- sort(ref.val)

sev

# Upper bound
Un <- 0.001

# Test
(sev[1] < Un)
```

# Bootstrap for distance-based criteria and Hessian method

```{r}
#######################################################################################
# Bootstrap
#######################################################################################

# Number of Boostrap samples 
B <- 1000

# Bootstrap MLEs
MLE.B <- matrix(0, ncol = length(MLE.PGWGH), nrow = B)
neB <- UKLB <- UHB <- DHB<- DKLB <- vector()
indHess <- vector()

for(i in 1:B){
  ind <- sample(1:n, replace = T)
  OPTB <- GHMLE(init = c(0,0,0, rep(0, ncol(xt)), rep(0,ncol(x))), times = times[ind], status = status[ind], 
           hstr = "GH", dist = "PGW", des = x[ind,], des_t = xt[ind,], method = "nlminb", maxit = 10000)
  MLE.B[i,] <- c(exp(OPTB$OPT$par[1:3]), OPTB$OPT$par[-c(1:3)])
  neB[i] <- n - 0.5*sum(!status[ind])
  
  UKLB[i] <- 0.5*k*M*length(MLE.B[i,])*log(neB[i])/neB[i]
  
  # Hessian/Eigenvalues
  HESSB  <- -hessian(OPTB$log_lik, x = OPTB$OPT$par)
  
  eigen.val <- eigen(HESSB)$values
  
  ref.val <- abs(as.vector(eigen.val))/max(abs(as.vector(eigen.val)))
  
  sev <- sort(ref.val)
  
  # Test
  indHess[i] <- as.numeric(sev[1] < Un)
}


######################################################################
# Bootstrap KL Divergence criterion
######################################################################

# Bootstrap KL Divergence

for(i in 1:B){
    if(MLE.B[i,3] < 0.01) DKLB[i] <- NA
    if(MLE.B[i,3] >= 0.01) DKLB[i] <- minKL(MLE.B[i,1:3], c(1,1))$min.KL
}

# Bootstrap probability of KL divergence criterion

indKLB <- (DKLB < UKLB) 

mean(indKLB, na.rm = TRUE)




######################################################################
# Bootstrap Hellinger distance criterion
######################################################################

# Hellinger Distance
for(i in 1:B){
DHB[i] <- minHell(MLE.B[i,1:3], c(1,1))$min.dHell

# Hellinger criterion
UHB[i] <- log( 1 - (1-2*kappa)^2)/(2*log(1-DHB[i]^2))
}

# Bootstrap probability of Hellinger distance criterion
indHB <- (neB < UHB)

mean(indHB)

######################################################################
# Bootstrap Hessian method
######################################################################


# Bootstrap probability of KL divergence criterion

mean(indHess) 
```


# Detecting Practical Non-Identifiability
```{r}
#######################################################################################
# Profile likelihood
#######################################################################################
p0 <- ncol(xt)
p1 <- ncol(x)

p <- length(MLE.PGWGH) # number of parameters
ML <- OPTPGWGH$OPT$objective # Maximum likelihood


# Profile likelihood function for parameter "ind"
prof.lik <- function(par1, ind){
  
  tempf <- function(par){
    tempv <- rep(0,p)
    tempv <- replace(x = tempv, c(1:p)[-ind] , par)
    tempv <- replace(x = tempv, ind , par1)
    out0 <- OPTPGWGH$log_lik(tempv)
    return(out0)
  } 
  
  out <-  -nlminb(OPTPGWGH$OPT$par[-ind]*0,tempf, control = list(iter.max = 10000))$objective + ML

  out2 <- ifelse(exp(out)<=1, exp(out), 0)
  return(out2)
}


# Profile likelihoods

# Profile likelihood of Parameter 1
prof1 <- Vectorize(function(par) prof.lik(log(par),1))
curve(prof1,0.01,7 , n = 500, lwd = 2, xlab = expression(sigma), ylab = "Profile Likelihood")

# Profile likelihood of Parameter 2
prof2 <- Vectorize(function(par) prof.lik(log(par),2))
curve(prof2,0.8,2 , n = 500, lwd = 2, xlab = expression(nu), ylab = "Profile Likelihood")

# Profile likelihood of Parameter 3
prof3 <- Vectorize(function(par) prof.lik(log(par),3))
curve(prof3,0.25,2.5 , n = 200, lwd = 2, xlab = expression(gamma), ylab = "Profile Likelihood")

# Profile likelihood of Parameter 4
prof4 <- Vectorize(function(par) prof.lik(par,4))
curve(prof4,-5,5, n = 500, lwd = 2, xlab = expression(alpha[1]), ylab = "Profile Likelihood", 
      ylim = c(0,1))

# Profile likelihood of Parameter 5
prof5 <- Vectorize(function(par) prof.lik(par,5))
curve(prof5,-2,2 , n = 200, lwd = 2, xlab = expression(beta[1]), ylab = "Profile Likelihood",
      ylim = c(0,1))


# Profile likelihood of Parameter 6
prof6 <- Vectorize(function(par) prof.lik(par,6))
curve(prof6,-1.1, -0.1 , n = 200, lwd = 2, xlab = expression(beta[2]), ylab = "Profile Likelihood")

# Profile likelihood of Parameter 7
prof7 <- Vectorize(function(par) prof.lik(par,7))
curve(prof7,0,0.65, n = 200, lwd = 2, xlab = expression(beta[3]), ylab = "Profile Likelihood")

```


# Alternative models
```{r}

################################################
# Exponentiated Weibull model
################################################


# Exponentiated Weibull

OPTEWGH <- GHMLE(init = OPTPGWGH$OPT$par*0.1, times = times, status = status, 
                 hstr = "GH", dist = "EW", des = x, des_t = xt, method = "nlminb", maxit = 10000)

# MLE EW-GH structure
MLE.EWGH <- c(exp(OPTEWGH$OPT$par[1:3]), OPTEWGH$OPT$par[-c(1:3)])
kable(MLE.EWGH, digits = 3)
AIC.EWGH <- 2*OPTEWGH$OPT$objective + 2*length(MLE.EWGH)
AIC.EWGH

################################################
# Weibull models (Simpler identifiable model)
################################################

# Optimisation
OPTW0 <- GHMLE(init = c(0,0), times = times, status = status, 
          hstr = "baseline", dist = "Weibull", method = "nlminb", maxit = 10000)

OPTWPH <- GHMLE(init = c(OPTW0$OPT$par*0, rep(0,ncol(x))), times = times, status = status, 
       hstr = "PH", dist = "Weibull", des = x, method = "nlminb", maxit = 10000)

OPTWAFT <- GHMLE(init = c(OPTW0$OPT$par*0,rep(0,ncol(x))), times = times, status = status, 
     hstr = "AFT", dist = "Weibull", des = x, method = "nlminb", maxit = 10000)

# MLE W no covariates
MLE.W0 <- exp(OPTW0$OPT$par[1:2])
kable(MLE.W0, digits = 3)

# MLE W-PH structure
MLE.WPH <- c(exp(OPTWPH$OPT$par[1:2]), OPTWPH$OPT$par[-c(1:2)])
kable(MLE.WPH, digits = 3)

# MLE W-AFT structure
MLE.WAFT <- c(exp(OPTWAFT$OPT$par[1:2]), OPTWAFT$OPT$par[-c(1:2)])
kable(MLE.WAFT, digits = 3)

# AIC
AIC.W0 <- 2*OPTW0$OPT$objective + 2*length(MLE.W0)
AIC.WAFT <- 2*OPTWAFT$OPT$objective + 2*length(MLE.WAFT)
AIC.WPH <- 2*OPTWPH$OPT$objective + 2*length(MLE.WPH)

# Best model: AFT-W
c(AIC.W0, AIC.WPH, AIC.WAFT)
```
